Methods paper for erosion

### 2017-10-24 docs/methods/erosion_methods.tex
worked on formulation and numerics.  Wrote a plan for numerical examples

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Pressure}
To compute the pressure at $\xx \in \overline{\Omega}$, we follow the
same procedure used for the shear stress.  We start by computing the
pressure of the double-layer potential for $\xx \in \Omega$ and add the
appropriate jump if $\xx \in \Gamma$. For $\xx \in \Omega$, the pressure
is
\begin{align}
  p^{D}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot \eeta(\yy) \,ds_\yy.
    \label{eqn:pressureDLP}
\end{align}
To compute the pressure at $\xx_0 \in \bd\Omega$, we consider the limit
as $\xx$ tends to $\xx_0$.  The limiting value is given
by~\cite{qua-bir2014a}
\begin{align}
  \lim_{\substack{\xx \rightarrow \xx_0 \\ \xx \in \Omega}}p^\DD(\xx) =  
    \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx_0).
  \label{eqn:pressureJump}
\end{align}
The pressure due to the Stokeslets is
\begin{align*}
  p^S(\xx_0) = \sum_{k=1}^{M}\frac{\rr \cdot \llambda_k}{2\pi\rho^2},
\end{align*}
and the pressure due to the rotlets is zero.  Therefore, the pressure is
\begin{align*}
  p(\xx) &= p^{\DD}(\xx) + p^{S}(\xx), \quad &&\xx \in \Omega,\\
  p(\xx) &= \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx) + 
              p^{S}(\xx), && \xx \in \bd\Omega.
\end{align*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Shear Stress}
We have decomposed the deformation tensor into the contributions due to
three different terms.  The contributions from the Stokeslets and the
rotlets can be directly evaluated since $\rr \neq 0$ for all $\xx \in
\gamma$.  However, the kernel of the deformation tensor due to the
double-layer potential~\eqref{eqn:shearStressDLP} 
\begin{align*}
  K_\ssigma(\xx,\yy)\eeta(\yy) &= \frac{1}{2\pi} \left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn)
    \right. \\ & \hspace{25pt}+ \left.
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right),
\end{align*}
has a singularity that is asymptotic to $\bigO(\rho^{-2})$ when $\yy
\rightarrow \xx$.  The singularity strength can be reduced since the
velocity corresponding to a constant density function is
constant~\cite{poz1992}.  Therefore, the deformation tensor of the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx) = \int_{\bd\Omega}K_\ssigma(\xx,\yy)
      (\eeta(\yy) - \eeta(\xx)) \, ds_{\yy}, \quad \xx \in \bd\Omega.
\end{align*}
This expression for the deformation tensor has a singularity that is
asymptotic to $\bigO(\rho^{-1})$ when $\yy \rightarrow \xx$.  Finally,
to achieve spectral accuracy, we apply odd-even
integration~\cite{sid-isr1988}.  That is, the deformation tensor due to
the double-layer potential at a discretization point $\xx_i$ on body
$\gamma_k$, with $i$ being odd, is approximated as
\begin{align}
  \ssigma^{\DD}(\xx_i) = \sum_{\xx_j \notin \gamma_k}
    K_\ssigma(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_j + 
  \sum_{\substack{\xx_j \in \gamma_k \\ j \: \mathtt{even}}}
    K_\ssigma(\xx_i,\xx_j) (\eeta(\xx_j) - \eeta(\xx_i)) \Delta s_j,
  \label{eqn:stressOddEven}
\end{align}
and a similar approximation is used when $i$ is even.  The deformation
tensor due to the Stokeslets and rotlets is given in
equation~\eqref{eqn:shearStressSR}.  Finally, the jump in the
deformation tensor~\eqref{eqn:shearStressJump} is added, and the
required arclength derivative is computed in Fourier space.  Then, the
shear stress can be computed by computing the deformation tensor in the
normal direction, and taking the inner product with the tangent
direction.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Pressure}
Similar to the shear stress, the singularity of the pressure
corresponding to the double-layer potential~\eqref{eqn:pressureDLP} is
too strong for the trapezoid rule to be directly applied.  Therefore, we
again subtract a constant density function and write the pressure as
\begin{align*}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx_0)) \,ds_\yy.
\end{align*}
Then, identical to how the shear stress is
computed~\eqref{eqn:stressOddEven}, odd-even integration is used to
evaluate this layer potential for pressure.  Finally, to add the jump
condition~\eqref{eqn:pressureJump}, Fourier differentiation is used to
compute the arclength derivative.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature: Off Boundary Calculations} 
\label{sec:off_boundary}
To analyze the flow, we compute quantities at target points that are in
the interior of the domain.  In particular, we develop numerical methods
to compute the velocity~\eqref{eqn:velocityDLP},
pressure~\eqref{eqn:pressureDLP}, vorticity~\eqref{eqn:vorticityDLP} and
resistivity~\eqref{eqn:resistivity}.

Similar to when the target points are on the boundary of the domain, all
the integrals, except for the resistivity, involve integrands that are
both periodic and smooth.  However, when the target points are close to
the boundary of the domain, the integrand becomes nearly-singular, and
these large derivatives result in a loss of accuracy that needs to be
rectified.  We will be adopting a near-singular integration scheme that
allows us to compute all quantities of interest sufficiently close to
$\bd\Omega$ for the results we are untested in.

@@

Computing the velocity, pressure, and vorticity at a point $\xx \in
\Omega$ requires evaluating the layer potential
\begin{align*}
  \phi(\xx) = \int_{\bd\Omega} K(\xx,\yy) \eeta(\yy) ds_\yy,
\end{align*}
where $\phi$ is either the velocity, pressure, or vorticity, and the
kernel $K$ is in
equations~\eqref{eqn:velocityDLP},~\eqref{eqn:vorticityDLP},
and~\eqref{eqn:pressureDLP}.  This integral is approximated with a
trapezoid rule, which can be easily done with the $N$ points where the
density function has been solved
\begin{align}
  \phi(\xx) \approx \sum_{k=1}^N q_k K(\xx,\yy_k),
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density
function.  Based on numerical experiments, the error of the $N$-point
trapezoid rule is acceptable for points $\xx$ that are more than 5
arclength spacings from $\bd\Omega$.

For points that are closer than 5 arclength spacings to $\bd\Omega$, we
use an upsampled trapezoid rule.  The required upsampling of the
geometry and the density function is done with a Fourier interpolant.
To minimize the number of discretization points that are used to
approximate the layer potential~\eqref{eqn:layerPotential}, we the
upsampling rate given in Table~\ref{tbl:upsampling}.  For points that
are closer than $5/16$ of an arclength from $\bd\Omega$, a value of 0 is
simply assigned to $\phi$.

In future work, such as sedimenting eroding bodies, bodies will come too
close to one another for upsampling to only used.  Therefore, we plan to
implement a near-singular integration scheme that uses either
near-singular interpolation~\cite{qua-bir2014a}, quadrature by
expansion~\cite{klo-bar-gre-one2013}, Barycentric
interpolation~\cite{bar-wu-vee2015}, or panel-based
quadrature~\cite{hel-oja2008a}.

\begin{table}[htpb]
\centering
\begin{tabular}{|c|ccccc|}
  \hline
  $d(\xx,\bd\Omega)$ &
  $[5h,\infty]$ &
  $[\frac{5}{2}h,5h]$ &
  $[\frac{5}{4}h,\frac{5}{2}h]$ & 
  $[\frac{5}{8}h,\frac{5}{4}h]$ &
  $[\frac{5}{16}h,\frac{5}{8}h]$ \\ [2ex]
  \hline
  Upsample Rate & 1 & 2 & 4 & 8 & 16 \\
  \hline
\end{tabular}
\caption{\label{tbl:upsampling}The upsampling rate used in our
near-singular integration scheme.  $d(\xx,\bd\Omega)$ is the distance
between a points $\xx \in \Omega$ and the boundary of the domain
$\bd\Omega$, and $h$ is an arclength spacing.}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Accelerating the Layer Potential Evaluations}
\label{sec:fmm}
To compute the velocity at an interior point $\xx \in \Omega$, the
integrand is periodic and smooth, so the trapezoid rule rule achieves
spectral accuracy.  Therefore, the velocity at a target points $\xx \in
\Omega$ is approximate by
\begin{align*}
  \uu(\xx) = \sum_{j=1}^N K(\xx,\xx_j) \eeta(\xx_j) \Delta s_j.
\end{align*}
If the velocity needs to be computed at $N_\tar$ points in $\Omega$,
where $N_\tar$ is large, the FMM is used to reduce the complexity from
$\bigO(N_\tar N)$ to $\bigO(N_\tar + N)$.

The dense linear system~\eqref{eqn:completed_BIE_discretized} is solved
iteratively with GMRES~\cite{saa-sch1986}.  The most expensive part is
computing the $N$ term sum for all target points.  A direct summation
would require $\bigO(N^2)$ operations, but this is reduced to $\bigO(N)$
by using the Fast Mutlipole method for the Stokes double-layer
potential~\cite{gre-rok1987,rac-???} \todo[inline]{sent email to Manas
asking about this}.  Since we are discretizing a second-kind integral
equation, the number of GMRES iterations is bounded independent of
$N$~\cite{cam-ips-kel-mey-xue1996}, and the overall complexity for
solving~\eqref{eqn:completed_BIE_discretized} is $\bigO(N)$.

In our implementation, the deformation tensor is computed with a direct
summation, and this requires $\bigO(N^2)$ operations.  Since this
calculation is only computed once per time step, while the velocity is
calculated at each GMRES iteration, the computational cost is
manageable.  However, in future work, to simulate erosion in a dense
packing of grains, a fast summation method, such as the
kernel-independent FMM~\cite{yin-bir-zor2004}, will be incorporated.

@@

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Pressure}
To compute the pressure at $\xx \in \overline{\Omega}$, we follow the
same procedure used for the shear stress.  We start by computing the
pressure of the double-layer potential for $\xx \in \Omega$ and add the
appropriate jump if $\xx \in \Gamma$. For $\xx \in \Omega$, the pressure
is
\begin{align}
  p^{D}(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot \eeta(\yy) \,ds_\yy.
    \label{eqn:pressureDLP}
\end{align}
To compute the pressure at $\xx_0 \in \bd\Omega$, we consider the limit
as $\xx$ tends to $\xx_0$.  The limiting value is given
by~\cite{qua-bir2014a}
\begin{align}
  \lim_{\substack{\xx \rightarrow \xx_0 \\ \xx \in \Omega}}p^\DD(\xx) =  
    \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx_0).
  \label{eqn:pressureJump}
\end{align}
The pressure due to the Stokeslets is
\begin{align}
  p^S(\xx_0) = \sum_{k=1}^{M}\frac{\rr \cdot \llambda_k}{2\pi\rho^2},
  \label{eqn:shearPressureSR}
\end{align}
and the pressure due to the rotlets is zero.  Therefore, the pressure is
\begin{align*}
  p(\xx) &= p^{\DD}(\xx) + p^{S}(\xx), \quad &&\xx \in \Omega,\\
  p(\xx) &= \pderiv{\eeta}{\ss} \cdot \ss + p^{\DD}(\xx) + 
              p^{S}(\xx), && \xx \in \bd\Omega.
\end{align*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Shear Stress}
We have decomposed the deformation tensor into the contributions due to
three different terms.  The contributions from the Stokeslets and the
rotlets can be directly evaluated since $\rr \neq 0$ for all $\xx \in
\gamma$.  However, the kernel of the deformation tensor due to the
double-layer potential~\eqref{eqn:shearStressDLP} 
\begin{align*}
  K_\ssigma(\xx,\yy)\eeta(\yy) &= \frac{1}{2\pi} \left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn)
    \right. \\ & \hspace{25pt}+ \left.
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right),
\end{align*}
has a singularity that is asymptotic to $\bigO(\rho^{-2})$ when $\yy
\rightarrow \xx$.  The singularity strength can be reduced since the
velocity corresponding to a constant density function is
constant~\cite{poz1992}.  Therefore, the deformation tensor of the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx) = \int_{\bd\Omega}K_\ssigma(\xx,\yy)
      (\eeta(\yy) - \eeta(\xx)) \, ds_{\yy}, \quad \xx \in \bd\Omega.
\end{align*}
This expression for the deformation tensor has a singularity that is
asymptotic to $\bigO(\rho^{-1})$ when $\yy \rightarrow \xx$.  Finally,
to achieve spectral accuracy, we apply odd-even
integration~\cite{sid-isr1988}.  That is, the deformation tensor due to
the double-layer potential at a discretization point $\xx_i$ on body
$\gamma_k$, with $i$ being odd, is approximated as
\begin{align}
  \ssigma^{\DD}(\xx_i) = \sum_{\xx_j \notin \gamma_k}
    K_\ssigma(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_j + 
  \sum_{\substack{\xx_j \in \gamma_k \\ j \: \mathtt{even}}}
    K_\ssigma(\xx_i,\xx_j) (\eeta(\xx_j) - \eeta(\xx_i)) \Delta s_j,
  \label{eqn:stressOddEven}
\end{align}
and a similar approximation is used when $i$ is even.  The deformation
tensor due to the Stokeslets and rotlets is given in
equation~\eqref{eqn:shearStressSR}.  Finally, the jump in the
deformation tensor~\eqref{eqn:shearStressJump} is added, and the
required arclength derivative is computed in Fourier space.  Then, the
shear stress can be computed by computing the deformation tensor in the
normal vector, and taking the inner product with the tangent vector.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Pressure}
Similar to the shear stress, the singularity of the pressure
corresponding to the double-layer potential~\eqref{eqn:pressureDLP} is
too strong for the trapezoid rule to be directly applied.  Therefore, we
again subtract a constant density function and write the pressure as
\begin{align*}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx)) \,ds_\yy, 
    \quad \xx \in \bd\Omega
\end{align*}
Then, identical to how the shear stress is
computed~\eqref{eqn:stressOddEven}, odd-even integration is used to
evaluate this layer potential for pressure.  The pressure due to the
Stokeslets are given in equation~\eqref{eqn:shearPressureSR}.  Finally,
to add the jump condition~\eqref{eqn:pressureJump}, Fourier
differentiation is used to compute the arclength derivative.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature: Off Boundary Calculations} 
\label{sec:off_boundary}
To analyze the flow, we compute quantities at target points that are in
the interior of the domain.  In particular, we develop numerical methods
to compute the velocity~\eqref{eqn:velocityDLP},
pressure~\eqref{eqn:pressureDLP}, vorticity~\eqref{eqn:vorticityDLP} and
resistivity~\eqref{eqn:resistivity}.  Since the target points are not on
the boundary of the domain, all the integrals, except for the
resistivity, involve integrands that are both periodic and smooth.
However, when the target points are close to the boundary of the domain,
the integrand becomes nearly-singular, and these large derivatives
result in a loss of accuracy that needs to be rectified.  We will be
adopting a near-singular integration scheme that allows us to compute
all quantities of interest sufficiently close to $\bd\Omega$ for the
quantities of interest.

@@

Computing the velocity, pressure, and vorticity at a point $\xx \in
\Omega$ requires evaluating the layer potential
\begin{align*}
  \phi(\xx) = \int_{\bd\Omega} K(\xx,\yy) \eeta(\yy) ds_\yy,
\end{align*}
where $\phi$ is either the velocity, pressure, or vorticity, and the
kernel $K$ is in
equations~\eqref{eqn:velocityDLP},~\eqref{eqn:vorticityDLP},
or~\eqref{eqn:pressureDLP}.  This integral is approximated with a
trapezoid rule, 
\begin{align}
  \phi(\xx) = \sum_{k=1}^N q_k K(\xx,\yy_k),
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density
function.  Note that this sum can only be computed once
equation~\eqref{eqn:completed_BIE} has been solved for the density
function $\eeta$.  Based on numerical experiments, the error of the
$N$-point trapezoid rule is acceptable for points $\xx$ that are more
than 5 arclength spacings from $\bd\Omega$.

For points that are closer than 5 arclength spacings to $\bd\Omega$, we
use an upsampled trapezoid rule.  The required upsampling of the
geometry and the density function is done with a Fourier interpolant.
To minimize the number of discretization points that are used to
approximate the layer potential~\eqref{eqn:layerPotential}, we use the
same upsampling rate for all target points that are in the same dyadic
interval (Table~\ref{tbl:upsampling}).  For points that are closer than
$5/16$ of an arclength from $\bd\Omega$, a value of 0 is simply assigned
to $\phi$.
\begin{table}[htpb]
\centering
\begin{tabular}{|c|ccccc|}
  \hline
  $d(\xx,\bd\Omega)$ &
  $[5h,\infty]$ &
  $[\frac{5}{2}h,5h]$ &
  $[\frac{5}{4}h,\frac{5}{2}h]$ & 
  $[\frac{5}{8}h,\frac{5}{4}h]$ &
  $[\frac{5}{16}h,\frac{5}{8}h]$ \\ [2ex]
  \hline
  Upsample Rate & 1 & 2 & 4 & 8 & 16 \\
  \hline
\end{tabular}
\caption{\label{tbl:upsampling}The upsampling rate used in our
near-singular integration scheme.  $d(\xx,\bd\Omega)$ is the distance
between a points $\xx \in \Omega$ and the boundary of the domain
$\bd\Omega$, and $h$ is an arclength spacing.}
\end{table}

In future work, such as sedimenting eroding bodies, bodies will come too
close to one another for upsampling to only used.  Therefore, we plan to
implement a near-singular integration scheme that uses either
near-singular interpolation~\cite{qua-bir2014a}, quadrature by
expansion~\cite{klo-bar-gre-one2013}, Barycentric
interpolation~\cite{bar-wu-vee2015}, or panel-based
quadrature~\cite{hel-oja2008a}.  In addition, even with accurate
quadrature, bodies can overlap because of discretization errors, and
this can be avoided with contact algorithms such as, for example, recent
work of Lu et al.~\cite{lu-rah-zor2017}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Accelerating the Layer Potential Evaluations}
\label{sec:fmm}
The bulk of the computational work is evaluating discretizations of
layer potentials~\eqref{eqn:layerPotential} at a large number of points.
If a direct summation method is used to evaluate these sums at $N_\tar$
points, this requires $\bigO(NN_\tar)$ operations.  The layer potential
that is evaluated most frequently is in the GMRES iteration where the
layer potential $\DD$ is evaluated to compute the velocity at all points
on the boundary must be evaluated~\eqref{eqn:DLP}.  To reduce the
computational cost of this calculation, we use the Fast Multipole Method
(FMM)~\cite{gre-rok1987, gre-gre-may1992} to reduce the complexity from
$\bigO(N^2)$ to $\bigO(N)$.  Since we are discretizing a second-kind
integral equation, the number of GMRES iterations is bounded independent
of $N$~\cite{cam-ips-kel-mey-xue1996}, and the overall complexity for
solving~\eqref{eqn:completed_BIE_discretized} is $\bigO(N)$.

In our implementation, the deformation tensor and vorticity are computed
with a direct summation, and this requires $\bigO(N^2)$ and $\bigO(N
N_\tar)$ operations, respectively.  Since this calculation is only
computed once per time step, while the velocity is calculated at each
GMRES iteration, the computational cost is manageable.  However, in
future work, to simulate erosion in a dense packing of grains, a fast
summation method, such as the kernel-independent
FMM~\cite{yin-bir-zor2004}, will be incorporated.

@@

Runs to do
\begin{itemize}
  \item Single body
  \begin{itemize}
    \item Aspect ratio and opening angle
    \item Drag and resistivity
    \item Anisotropy
    \item Vorticity
  \end{itemize}
  \item $\bigO(5)$ bodies
  \begin{itemize}
    \item Drag and resistivity
    \item Anisotropy
    \item Long channels between flat faces
    \item Vorticity
  \end{itemize}
  \item $\bigO(50)$ bodies
  \begin{itemize}
    \item Drag and resistivity
    \item Anisotropy
    \item Vorticity
  \end{itemize}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{\bf Acknowledgments} The authors would like to thank Manas Rachh
for supplying the FMM for the Stokes double-layer potential.
